A performance assessment of evolutionary algorithms in a volunteer computing system

### 2016-03-14 PPSN2016/paper.Rnw
REST & Experiments sction

The architecture of the {\sf NodIO} system is divided in two tiers:\begin{enumerate}
\item A REST (representational state transfer) server that responds to
  HTTP requests made by the browsers. These HTTP requests use JSON for
  encoding information, in this case chromosomes and associates
  information send by the clients and return requests made by the
  server.  The server has the capability to
  run a single experiment, storing the incoming chromosomes in a cache data structure
  that is reset when the solution is found. This data has a finite
  size that erases the oldest chromosomes once it has filled to
  capacity.  
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays the fitness and
  other graphs, some
  additional links, and information on the experiment. This code runs
  an evolutionary algorithm {\em island} that starts with a random
  population, then after every 100 generations, it sends the best individual
  back to the server (via a {\tt PUT} request), and then requests a random
  individual back from the server (via a {\tt GET} request). We have
  kept the number of generations between migrations fixed since it is
  a way of finding out how much real work every client is doing. In
  the version of the server used in this paper, we implemented a
  policy by which the combination IP-fitness value could only be
  inserted once into the cache. That means that if a particular IP
  sends the fitness 99 into the server, it will be accepted only
  once. The main intention of this policy was to avoid {\em
    overlapping} of contributions, with clients still running an {\em
    old} run of the experiment contributing after a new one had
  started. This did not completely avoid these contributions: if the
  clients of the {\em old} run increased their new fitness, it would
  still be allowed. 
  % As I remember we kept the run_number so IPs with old runs
  % did not interfere with newer ones. But you are saying: what if they are super slow 
  % and always PUT when the run is already over?, we want to receive 
  % those individuals? 
  % "That means that if a particular IP
  % sends the fitness 99 into the server, it will be accepted only
  % once."  
  % once per run?
% JJ - once per run or experiment. But we didn't keep that run_number
% to avoid interference, just to keep track
  %  but different individuals could have the same fitness, are we limiting the search? 
% We are actually checking for the same individual, not only the fitness
  %  maybe we could have multiple versions of the pool a la MVCC  :)
% Let's leave that for the future...- JJ
  % I think if we are just trying to limit overlapping of the experiments
  % then PUTs should send the run's seq_number and completly remove it.  
% The client does not keep track of the seq number. Maybe we should do
% just that, but sending back and forth the seq number implies
% refactoring. - JJ
\end{enumerate}

@@

NodIO needs a fitness function to work with; in this case the classical Trap 
function \cite{Ackley1987} has been used; this function runs in the
browser, but the server needs to know when the solution has been
found, so a termination condition is also included as a configuration
option in the server. In all cases we are using 50 Traps, a problem
that is difficult enough to need the intervention of several users for
an extended amount of minutes.

For every new individual sent by the clients to the cache, several
% We have not presented the "cache",  only the server and
% ..storing the incoming chromosomes in a data structure
% Im changing over there to cache
pieces of meta data were stored: a time stamp, the client IP, the chromosome and
value, the cache size in that particular moment, and also if that
individual actually updated the cache or not.

@@

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. In fact, the NodIO system can be deployed to any
system as long as it can run node.js and access to the filesystem, for
log files, is available. All these logs are available under
rrequest. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\label{fig:zipf}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP order) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that is slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

The baseline, called {\em base} in the Table, uses simply the
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. The fact that these users were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache and labeled ``How am I contributing''. The user could
then realize if his contribution was being 0 and thus do something
about it; these are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if evolution got unstuck, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently initiated. This batch of runs is labelled {\tt
  reboot}. However, we found that this size of the cache was actually
skipped by some slow clients or too fast evolutions, so that in the
last case, labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage engagement by the users and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.

>>>>

The architecture of the {\sf NodIO} system is divided in two tiers:\begin{enumerate}
\item A REST (representational state transfer) server that responds to
  HTTP requests made by the browsers. These HTTP requests use JSON for information 
  encoding, in this case chromosomes and related
  information sent by the clients and responses returned by the
  server.  The server has the capability to
  run a single experiment, storing the incoming chromosomes in a cache data structure
  that is reset when the solution is found. This cache has a finite
  size that erases the oldest chromosomes once it has filled to
  capacity.  
\item A client that includes the evolutionary algorithm as
  JavaScript code embedded in a web page that displays the fitness and
  other graphs, some
  additional links, and information on the experiment. This code runs
  an evolutionary algorithm {\em island} that starts with a random
  population, then after every 100 generations, it sends the best individual
  back to the server (via a {\tt PUT} request), and then requests a random
  individual back from the server (via a {\tt GET} request). We have
  kept the number of generations between migrations fixed since it is
  a way of finding out how much real work every client is doing. In
  the version of the server used in this paper, we implemented a
  policy by which the combination IP-fitness chromosome could only be
  inserted once into the cache. That means that if a particular IP
  sends the fitness 99 into the server, it will be accepted only
  once. The main intention of this policy was to avoid {\em
    overlapping} of contributions, with clients still running an {\em
    old} run of the experiment contributing after a new one had
  started. This did not completely avoid these contributions: if the
  clients of the {\em old} run increased their new fitness, it would
  still be allowed. 
  % As I remember we kept the run_number so IPs with old runs
  % did not interfere with newer ones. But you are saying: what if they are super slow 
  % and always PUT when the run is already over?, we want to receive 
  % those individuals? 
  % "That means that if a particular IP
  % sends the fitness 99 into the server, it will be accepted only
  % once."  
  % once per run?
% JJ - once per run or experiment. But we didn't keep that run_number
% to avoid interference, just to keep track
  %  but different individuals could have the same fitness, are we limiting the search? 
% We are actually checking for the same individual, not only the fitness
  %  maybe we could have multiple versions of the pool a la MVCC  :)
% Let's leave that for the future...- JJ
  % I think if we are just trying to limit overlapping of the experiments
  % then PUTs should send the run's seq_number and completly remove it.  
% The client does not keep track of the seq number. Maybe we should do
% just that, but sending back and forth the seq number implies
% refactoring. - JJ
\end{enumerate}

@@

NodIO needs a fitness function to work with; in this case the classical Trap 
function \cite{Ackley1987} has been used; this function runs in the
browser, but the server needs to know when the solution has been
found, so a termination condition is also included in the server, as a configuration
option. In all cases we are using 50 Traps, a problem
that is difficult enough to need the intervention of several users for
an extended amount of minutes.

For every new individual sent by the clients to the cache, several
% We have not presented the "cache",  only the server and
% ..storing the incoming chromosomes in a data structure
% Im changing over there to cache
pieces of meta data were stored: a time stamp, the client IP, the chromosome and
fitness value, the cache size in that particular moment, and also if that
individual actually updated the cache or not.

@@

graph2 & 1173561.5 & 6 & 4 & 761.5 & 22.0 & 0.0520509 & 36 & 73 & 129.04545\\
\hline
reboot & 796214.0 & 9 & 6 & 646.0 & 67.0 & 0.0955121 & 45 & 168 & 75.88889\\
\hline
cachereboot & 392917.0 & 12 & 12 & 576.0 & 79.0 & 0.1575092 & 55 & 225 & 43.28571\\
\hline
\end{tabular}
%% \end{tabular}
%% Initially generated with this code, 
% <<experiments, cache=FALSE,echo=FALSE>>=
%experiment.aggregates$Experiment <- c("base","graph1","graph2","reboot","cachereboot")
%kable(experiment.aggregates)
%@ %def 
\end{center}
\end{table}
%
%
%
Since the point of using volunteer computing system is to cut costs,
all experiments were set up using the OpenShift
PaaS, which provides a free tier, making the whole experiment cost
equal to \$0.00. In fact, the NodIO system can be deployed to any
system as long as it can run node.js and has access to the filesystem, for
log files, if available. All these logs are available upon
request. Processed files are available in the same repository that
hosts this paper, as well as the scripts needed to process them.

%
%
\begin{figure}[htbp]
\centering
<<zipf, cache=FALSE,echo=FALSE,fig.height=3>>=
ggplot(ips.data,aes(x=idx,y=PUTs,group=Experiment,color=Experiment))+geom_point()+scale_y_log10()+ scale_colour_discrete(labels=c('base','graph1','graph2','reboot','cachereboot'))+scale_x_log10()
@ %def graph for IPs and puts
%
\caption{Ranked and normalized number of contributions (PUTs) per IP,
  plotted in a log $y$ scale.\label{fig:zipf}}
\label{fig:zipf}
\end{figure}
%
Every run had slightly different conditions, although in some cases it
was just a server reboot and a new round of announcements. Each run
batch included at least 30 completions, that is, evolution until the
solution was found. The number of runs was controlled by polling an
URL that indicated the number of runs made in the present batch so
far. Every volunteer contributed a certain number of individuals
(represented by the PUT HTTP request) to the server. The ranked number
of contributions per unique IP is shown in Figure \ref{fig:zipf} in a
log-log scale. Besides following the usual Zipf's law found in our
previous work and having a very similar appearance for all 5 sets, it
is interesting to see how some sets have a number of contributions per
IP that are slightly superior to the rest, specially so in {\tt
  graph2}. The last two sets, {\tt reboot} and {\tt cachereboot}, also
show a regime in the middle of the graph that shows that some users
are contributing more than in other similar sets. In general, these
differences in behavior are due more to the specific users that show
up in these experiments than to general laws, but it should also be
noted that those behaviors can be encouraged by design.

@@

The baseline, called {\em base} in the Table, uses the simple
mechanism of dropping contributions from clients if they are repeated
to avoid overlap. Clients still get a random chromosome from the
server, so they can in fact proceed with the algorithm and even finish
it. The fact that these users were not contributing to the pool was
conveyed by inserting a graph in the client that represented the size
of the cache and labeled ``How am I contributing''. The user could
then realize if his contribution was being 0 and thus do something
about it; these are labeled {\tt graph1} and 2. However, a stuck
client could still add to the pool if evolution got unstuck, so we
added a mechanism for rebooting the client, that is, reloading the
page, if the cache size was found to be less than 1, which indicated a
run recently initiated. This batch of runs is labelled {\tt
  reboot}. However, we found that this size of the cache was actually
skipped by some slow clients or too fast evolutions, so that in the
last case, labeled {\tt cachecrash}, clients rebooted if the cache
collapsed by more than a certain length, indicating a cache that had
been voided and was filling again. Our objective was twofold: to
encourage engagement by the users and also to increase the involvement
of every client in the common experiment, eliminating at the same time
overlaps between runs.
